Erosion of somatic tissue identity with loss of the X-linked intellectual disability factor KDM5C

### 2024-09-02 submission/manuscript.Rmd
discusssion edits

Many germline-specific genes are suppressed by the transcription factor heterodimers E2F6/DP1 and MGA/MAX, which respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @stielowMGAL3MBTL2E2F62018; @dahletE2F6InitiatesStable2021]. To elucidate if KDM5C is recruited to germline genes by a similar mechanism, we used HOMER[@heinzSimpleCombinationsLineageDetermining2010] to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters (TSS &#177; 500 bp, q-value < 0.1). MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5G). One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 5H). KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, RFX2 q-value < 0.0001, RFX5 q-value < 0.0001) (Figure 5I, Supplementary figure 1D). RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009] and among them is RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. Interestingly, RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5J), however it is also not a direct target of KDM5C (Supplementary figure 1E). Thus, RFX2 is a candidate transcription factor for driving the ectopic expression of KDM5C-unbound germline genes in _Kdm5c_-KO cells.

@@

**make edits**
Although eggs and sperm employ the same transcriptional programs for shared functions, e.g. PGC formation,  meiosis, and genome defense, some germline genes are sex specific. We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of derepression is independent of a given germline gene's sex. However, organismal sex did greatly influence the degree of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males. The lack of X-linked gene enrichment in females suggests that this greater sensitivity to germline gene misexpress is not due to XCI defects previously reported in _Kdm5c_-KO females[@samantaActivationXistEvolutionarily2022]. Sex differences in germline gene suppression may be a consequence of females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males. Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to identify the biological implications and underlying mechanisms.

@@

- In somatic cells, germline genes are highly methylated at promoter CGIs, which are typically unmethylated for other classes of genes.
- The classic model for germline gene silencing is accumulation of CpGme at CGIs within their promoter, which is facilitated by E2F6/DP1 and MAX/MGA binding sites
- KDM5C does not have sequence specific binding, associates with PRC1.6 but likely not the major mechanisms of recruitment 
- KDM5C highly enriched at meiotic, promoter CGI germline genes
- A plausible mechanism for KDM5C-mediated germline gene silencing is KDM5C's removal of H3K4me2/3 at germline gene CGIs[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given that (H3K4me2/3) oppose DNMTs and primarily do no colocalize with CpGme.
- Indeed we observed H3K4me2/3 accumulation at germline TSS and impaired CGI methylation at select germline genes. Loss of KDM5C strongly impairs CpGme at a subset of CGI-containing germline genes.
- However, Emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020].
- KDM5C's catalytic activity was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. In EpiLCs, Dazl still gained CpGme even with elevated H3K4me2. This suggests other redundant silencing mechanisms to ensure eventual CpGme of dazl, however other germline CGIs require KDM5C to ensure sufficient methylation
- Furthermore, our results indicates a histone modifier's catalytic and non-catlytic roles can change depending upon the locus and developmental stage.

- It is yet unknown how KDM5C is recruited to germline gene promoters, given that KDM5C itself does not contain domains for sequence-specific binding[@iwaseXLinkedMentalRetardation2007]. In HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023], KDM5C assocaiates with members of the polycomb repressive complex 1.6 (PRC1.6), which is recruited to germline gene promoters through E2F6/DP1 and MGA/MAX heterodimers[@endohPCGF6PRC1SuppressesPremature2017; @stielowMGAL3MBTL2E2F62018]. While MAX and E2F6 motifs were enriched at KDM5C-bound germline genes in EpiLCs, only about one-third of promoters contained their consesus sequence.

- Our list of germline genes
- We found less than 30% of germline-enriched genes have CGIs, yet the majority of CGI-free germline genes still gained significant CpGme around their TSS. 
	- CGI-containing = meiotic
- Loss of KDM5C binding at a subset of CGI-containing germline genes in EpiLCs, e.g. D1Pas1, Naa11, strongly impairs CpGme, resulting in their ectopic transcription in both EpiLCs and in the mature brain.
- Our germline gene list enabled us to identify germline-enriched genes activated independent of

- This analysis revealed some germline genes can become active in _Kdm5c_-KO cells independent of KDM5C binding to their promoters. 
- Germline genes unbound by KDM5C primarily late-stage spermatogenesis genes that lacked CGIs and significantly enriched for RFX transcription factors, including RFX2, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016].

Another notable KDM5C-unbound DEG is _Stra8_, a meiotic transcription factor that is turned on in germ cells with retionic acid signaling and DAZL expression[@endoPeriodicRetinoicAcid2015; @bowlesRetinoidSignalingDetermines2006]. Misexpression of _Dazl_ and _Rfx2_ and their downstream targets in _Kdm5c_-KO cells suggests that once activated, ectopic germline genes can then turn on other aberrant germline programs to loosely mimic germ cell development.

- KDM5C loss also activates RFX2 transctiption factors whose binding sites are enriched at non-CGI spermiogenesis genes.

- CGIs are resistant
	- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5001583/
	- https://pubmed.ncbi.nlm.nih.gov/14504655/

- Our results indicate KDM5C's removal of H3K4me2/3 may be necessary at a subset of CGI-containing germline genes, given they fail to accumulate CpGme during EpiLC differentiation. These CpGme-sensitive germline genes, such as _D1Pas1_ and _Naa11_, are also de-repressed in the mature _Kdm5c_-KO brain and consistently dysregulated in multiple _Kdm5c_-KO datasets[@scandagliaLossKdm5cCauses2017].

- H3K4me3 and CpGme typically do not colocalize
	- https://pubmed.ncbi.nlm.nih.gov/17334365/
	- https://www.nature.com/articles/s41594-017-0013-5
	- https://pubmed.ncbi.nlm.nih.gov/23664763/
	- https://www.ncbi.nlm.nih.gov/pmc/articles/PMC2718496/
	- https://pubmed.ncbi.nlm.nih.gov/19581485/

- Contrastingly, many germline genes had only a modest reduction in CpGme, including _Dazl_ even though its TSS accumulated substantial H3K4me2. 
- Multiple regulators repress dazl, dazl is not expressed in the mature _Kdm5c_-KO brain
	E2F6/MGA/MAX
		- Loss of many chromatin regulators affects CpGme (Multiple, non-redundant silencing mechansisms) - seems to be the crux point of germline gene silencing
		-  Because CGIs are typically resistant to CpGme (accurate?), germline CGIs may require a highly repressive histone landscape to recruit sufficient DNMTs to these loci

-

- Combine germline gene list, CGI, RFX information together. 
	- Benefit of the list is finding CGI difference and kdm5c indirect mechanisms of de-repression
    - List of genes enables us to propose this model
    - CGI positive meiosis/germline formation regulators turned on → these turn on CGI-negative late stage regulators → These promote downstream dysregulation long term

- CGIs highly determine KDM5C recruitment to germline gene promoters.
		- Other studies on germline gene repressors have shown they are important for CGIme, unclear if they participate in non-CGI TSS CpGme

>>>>

Many germline-specific genes are suppressed by the polycomb repressive complex 1.6 (PRC1.6), which contains transcription factor heterodimers E2F6/DP1 and MGA/MAX that respectively bind E2F and E-box motifs[@hurlinMgaDualspecificityTranscription1999; @pohlersRoleE2F6Restriction2005; @velascoDnmt3bRecruitmentE2F62010; @suzukiLossMAXResults2016; @endohPCGF6PRC1SuppressesPremature2017 @stielowMGAL3MBTL2E2F62018; @tatsumiDNMTsSETDB1Function2018; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. PRC1.6 members have been proposed to recruit KDM5C to germline gene promoters, given their association in HeLa cells and ESCs[@tahilianiHistoneH3K4Demethylase2007; @guptaGenomewideScreenReveals2023] and that KDM5C itself does not contain sequence-specific binding domains[@iwaseXLinkedMentalRetardation2007]. To elucidate how KDM5C is recruited to germline genes, we used HOMER[@heinzSimpleCombinationsLineageDetermining2010] to identify transcription factor motifs enriched at KDM5C-bound or unbound germline gene promoters (TSS &#177; 500 bp, q-value < 0.1). MAX and E2F6 binding sites were significantly enriched at germline genes bound by KDM5C in EpiLCs, but not at germline genes unbound by KDM5C (MAX q-value: 0.0068, E2F6 q-value: 0.0673, E2F q-value: 0.0917) (Figure 5G). One third of KDM5C-bound promoters contained the consensus sequence for either E2F6 (E2F, 5'-TCCCGC-3'), MGA (E-box, 5'-CACGTG-3'), or both, but only 17% of KDM5C-unbound genes contained these motifs (Figure 5H). KDM5C-unbound germline genes were intstead enriched for multiple RFX transcription factor binding sites (RFX q-value < 0.0001, RFX2 q-value < 0.0001, RFX5 q-value < 0.0001) (Figure 5I, Supplementary figure 1D). RFX transcription factors bind X-box motifs[@gajiwalaStructureWingedhelixProtein2000] to promote ciliogenesis[@swobodaRFXTypeTranscriptionFactor2000; @ashiqueRfx4TranscriptionFactor2009] and among them is RFX2, a central regulator of post-meiotic spermatogeneis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. Interestingly, RFX2 mRNA is derepressed in _Kdm5c_-KO EpiLCs (Figure 5J), however it is also not a direct target of KDM5C (Supplementary figure 1E). Thus, RFX2 is a candidate transcription factor for driving the ectopic expression of KDM5C-unbound germline genes in _Kdm5c_-KO cells.

@@

Although eggs and sperm employ the same transcriptional programs for shared functions, e.g. PGC formation,  meiosis, and genome defense, some germline genes are sex specific. We found _Kdm5c_ mutant males and females expressed both sperm and egg-biased genes, indicating the mechanism of derepression is independent of a given germline gene's sex. However, organismal sex did greatly influence the degree of germline gene dysregulation, as female _Kdm5c_-KO EpiLCs had over double the number of germline-enriched DEGs compared to males. The lack of X-linked gene enrichment in females suggests that this greater sensitivity to germline gene misexpress is not due to XCI defects previously reported in _Kdm5c_-KO females[@samantaActivationXistEvolutionarily2022]. Sex differences in germline gene suppression may be a consequence of females having a higher dose of KDM5C than males, due to its escape from XCI[@agulnikNovelGeneWidely1994; @carrelTissueLineagespecificVariation1996; @sheardownMouseSmcxGene1996; @xuSexSpecificExpressionXLinked2008]. Intriguingly, females with heterozygous loss of _Kdm5c_ also had over double the number of germline DEGs than hemizygous knockout males, even though their level of KDM5C should be roughly equivalent to that of wild-type males. Altogether, these results suggests germline gene silencing mechanims differ between males and females, which warrants further study to identify the biological implications and underlying mechanisms.

@@

While promoter CGIs are typically resistant to CpGme[@longProtectionCpGIslands2016], CGIs at germline promoters are hightly methylated in somatic cells[@weberDistributionSilencingPotential2007]. In EpiLCs, loss of KDM5C binding at a subset of germline gene promoters, e.g. _D1Pas1_ and _Naa11_, strongly impaired CGI methylation, which resulted in their long-lasting de-repression into adulthood. Removal of H3K4me2/3 at CGIs is a plausible mechanism for KDM5C-mediated germline gene suppression[@scandagliaLossKdm5cCauses2017; @aladhamiSystematicIdentificationFactors2023], given H3K4me2/3 primarily do not colocalize with CpGme[@weberDistributionSilencingPotential2007] and can oppose DNMT3 activity[@otaniStructuralBasisRecognition2009; @guoStructuralInsightAutoinhibition2015]. However, emerging work indicates many histone-modifying enzymes have non-cataltyic functions that influnce gene expression, sometimes even more potently than their catalytic roles[@aubertUnexpectedNoncatalyticRoles2019; @morganReevaluatingRolesHistonemodifying2020]. Indeed, KDM5C's catalytic activity was recently found to be dispensible for repressing _Dazl_ in ESCs [@guptaGenomewideScreenReveals2023]. In our study, _Dazl_'s promoter still gained CpGme in _Kdm5c_-KO exEpiLCs, even with elevated H3K4me2. _Dazl_ and a few other germline gene CGIs use multiple repressive mechanisms to facilitate CpGme[@pohlersRoleE2F6Restriction2005; @endohPCGF6PRC1SuppressesPremature2017; @dahletE2F6InitiatesStable2021; @mochizukiRepressionGermlineGenes2021]. Together, this suggests alternative silencing mechansims are sufficient to recruit DNMT3s to some germline CGIs, while others may require KDM5C-mediated H3K4me removal to overcome promoter CGI escape from CpGme. Furthermore, these results indicate the requirement for catalytic activity for the same class of genes can change depending upon the locus and developmental stage.

By generating a comprehensive list of mouse germline-enriched genes, we were able to reveal distinct repressive mechanisms governing early versus late-stage germline developmental programs. In EpiLCs, KDM5C was highly enriched at germline promoters containing CGIs, many of which had roles in early germ cell formation and meiosis. However, over 70% of germline-enriched gene promoters lacked CGIs, including the many KDM5C-unbound germline genes that were de-repressed in _Kdm5c_-KO cells. CGI-free, KDM5C-unbound germline genes were primarily late-stage spermatogenesis genes and significantly enriched for RFX2 binding sites, a central regulator of spermiogenesis[@kistlerRFX2MajorTranscriptional2015; @wuTranscriptionFactorRFX22016]. These data suggest that once activated during early embryogenesis, drivers of germline identity like _Rfx2_ and _Dazl_ turn on downstream germline programs that can loosely mimic germ cell development, ultimately culminating in the expression of spermiogenesis genes in the adult _Kdm5c_-KO brain. Therefore, we propose KDM5C is recruited to germline genes that shape germ cell formation via their promoter CGIs to act as break against runaway activation of germline-specific programs. Therefore, we propose KDM5C is recruited via promoter CGIs to genes that shape germ cell formation and acts as break against runaway activation of germline-specific programs.
